Particle-in-cell simulation of ion and electron acceleration in trans-relativistic shocks.

### 2017-11-17 papers/Physica2017/Romansky_Physica2017.tex
paper

\author{V I Romansky$^{1,2}$, A M Bykov$^1$, S M Osipov$^1$ and P E Gladilin$^1$}

@@

\begin{abstract}
The problem of Cosmic Ray (CR) origin and acceleration is a widely discussed problem of
modern astrophysics. It is considered that most likely CR sources are 
shock waves in supernova remnants or jets. One of the methods to study particle injection and acceleration by shock
waves is a particle-in-cell computer simulation. In this paper we present results of
simulations of relativistic shock waves with low lorenz-factor, based on our self-developed implicit particle-in-cell code. 
\end{abstract}
\input{intro.tex}
\input{shock.tex}
\input{conclusions.tex}
\input{biblio.tex}
\end{document}

>>>>

\author{V I Romansky$^{1,2}$, A M Bykov$^1$ and S M Osipov$^1$}

@@

\begin{abstract}
The problem of Cosmic Ray (CR) origin and acceleration is a widely discussed problem of
modern astrophysics. It is considered that most likely CR sources are 
shock waves in supernova remnants or jets. One of the methods to study particle injection and acceleration by shock
waves is a particle-in-cell computer simulation. In this paper we present results of
simulations of relativistic shock waves with low lorentz-factor, based on our self-developed implicit particle-in-cell code. 
\end{abstract}
\input{intro.tex}
\input{shock.tex}
\input{conclusions.tex}
\input{biblio.tex}
\end{document}

### 2017-11-17 papers/Physica2017/conclusions.tex
paper

\section{Conclusions}
In this paper we presented the developed 3D implicit particle-in-cell code, which can be applied for various problems in the collisionless plasma. We have tested it and obtained results, that are consistent with theoretical predictions. We studied the influence of admixture of heavy ions on the evolution of the shock wave and simulations have shown that the presence of heavy ions increases the number of accelerated particles. Further research is necessary in order to study this phenomena in detail.
\ack
A M Bykov, P E  Gladilin and S M  Osipov have developed the implicit fields solver module, which provides conservation of the divergenceless magnetic field. 
V I Romansky acknowledges support from RSF grant 16-12-10519 which was used to develop the
particle mover, parallelization and perform the testing of the code.

The results of the work were obtained using computational resources of Peter the Great Saint-Petersburg Polytechnic University Supercomputing Center (http://www.spbstu.ru).

>>>>

\section{Conclusions}
In this paper we studied trans-relativistic shock wave using our implicit particle-in-cell code. Simulations show, that particles injection in acceleration process highly depends on angle between flow bulk velocity and magnetic field. Also injection of protons is more effective than electrons for angles less then critical, but there is regime of more effective electrons acceleration when angle is close to critical. Further research is necessary in order to study this phenomena in detail.
\ack
V I Romansky acknowledges support from RSF grant 16-12-10519.
The results of the work were obtained using computational resources of Peter the Great Saint-Petersburg Polytechnic University Supercomputing Center (http://www.spbstu.ru).

### 2017-11-17 papers/Physica2017/intro.tex
paper

\section{Introduction}
One of the possible sources of high-energy cosmic rays are shock waves in astrophysical collisionless plasma due to the diffusive shock acceleration (DSA) mechanism, see \cite{Bell1978}, \cite{Blandford1978}. The particle acceleration on the shock front is sophisticated and nonlinear problem, because cosmic rays have strongly influence on the shock wave. The pressure of cosmic rays is comparable with pressure of background plasma and significatly modifies the front (see for example \cite{Bykov2014}). Also there are a instabilities in anisotropic plasma, which lead to strong field amplification and the observations of the synchrotron emission from supernova remnants (SNRs) show, that the magnetic field there is about 100 times stronger then the interstellar field (see for example \cite{Berezhko2003},\cite{Uchiyama2007}). One of the most powerful methods to explore processes in collisionless plasma is the Particle-in-Cell simulation.

>>>>

\section{Introduction}
One of the possible sources of high-energy cosmic rays are shock waves in astrophysical collisionless plasma due to the diffusive shock acceleration (DSA) mechanism, see \cite{Bell1978}, \cite{Blandford1978}. The particle acceleration on the shock front is sophisticated and nonlinear problem, because cosmic rays have have strong influence on the shock wave. The pressure of cosmic rays is comparable with pressure of background plasma and significatly modifies the front (see for example \cite{Bykov2014}). Also there are instabilities in anisotropic plasma, which lead to strong field amplification and the observations of the synchrotron emission from supernova remnants (SNRs) show, that the magnetic field there is about 100 times stronger then the interstellar field (see for example \cite{Berezhko2003},\cite{Uchiyama2007}). One of the most powerful methods to explore processes in collisionless plasma is the Particle-in-Cell simulation.

### 2017-11-17 papers/Physica2017/shock.tex
paper

\section{Shock wave simulation}
We studied the evolution of trans-relativistic shocks and spectum of accelerated particles. We chose relastivistic shocks with low lorenz-factor because the maximum energy of produced cosmic rays increases with shock wave lorenz-factor, but the efficiency of acceleration decreases, because it is more difficult for particle to cross fast moving front many times \cite{Ellison2013}. So we assume that intermediate case of trans-relativistic shocks provides the most effective acceleration.

@@

In simulation setup the homogeneous plasma flows in through the right boundary
and collides with the reflecting superconducting wall on the left boundary, causing the shock wave

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.86\textwidth]{fig/protons.eps} 
	\caption{Distribution of protons.}
	\label{protons}
\end{figure}
\begin{figure}[h!]
	\centering
	\includegraphics[width=0.86\textwidth]{fig/electrons.eps} 
	\caption{Distribution of electrons.}
	\label{electrons}
\end{figure}

Simulations are one-dimensional and have following parameters: initial flow lorenz-factor $\gamma = 1.5$, number densities $n_e = 10^{-4} \rm{cm}^{-3}$, $n_p = 10^{-4} \rm{cm}^{-3}$ , temperature $5\cdot10^8 \rm{K}$, magnetic field $B = 10^{-4} \rm{G}$, the full size of the box $L = 2\cdot10^{12} \rm{cm}$, the number of cells $N=2\cdot10^4$. Electron mass is reduced to $m_e = \frac{m_p}{20}$. The full time of simulation is $T = 5000 {\omega_p}^{-1}$. $\theta$ is angle between flow velocity and magnetic field. We present results of particle spectrum in several simulations with different $\theta$.

One can see from Figures \ref{protons} and \ref{electrons} that spectrum of accelerated particles strongly depends on angle theta. If $\theta$ is less then critical value, defined by equation $c\cdot cos(\theta_{crit})=v_{shock}$, where all values are measured in upstream rest frame, particles can escape from the front and cross it several times to gain more energy. It is difficult to evaluate critical angle apriori, because we don't know compression relation of shock wave which will be created in simulation, but estimation for case of strong wave (comression relation $\sigma=4^{\circ}$) gets $\theta_{crit}=45$ in downstream frame. And it is consistent with results of our simulations.

>>>>

\section{Shock wave simulation}
We studied the evolution of trans-relativistic shocks and spectum of accelerated particles. We chose relastivistic shocks with low lorentz-factor because the maximum energy of produced cosmic rays increases with shock wave lorentz-factor, but the efficiency of acceleration decreases, because it is more difficult for particle to cross fast moving front many times \cite{Ellison2013}. So we assume that intermediate case of trans-relativistic shocks provides the most effective acceleration.

@@

In simulation setup the homogeneous plasma flows in through the right boundary
and collides with the reflecting superconducting wall on the left boundary, causing the shock wave. It is common way to initialize shock wave, because initialization using Rankine–Hugoniot conditions does not take in account microscopic distribution and shock wave, created that way, will probably fall apart for several discontinuities.

  Simulations are one-dimensional and have following parameters: initial flow lorentz-factor $\gamma = 1.5$, number densities $n_e = 10^{-4} \rm{cm}^{-3}$, $n_p = 10^{-4} \rm{cm}^{-3}$ , temperature $5\cdot10^8 \rm{K}$, magnetic field $B = 10^{-4} \rm{G}$, the full size of the box $L = 2\cdot10^{12} \rm{cm}$, the number of cells $N=2\cdot10^4$. Electron mass is reduced to $m_e = \frac{m_p}{20}$. The full time of simulation is $T = 5000 {\omega_p}^{-1}$. $\theta$ is angle between flow velocity and magnetic field. We present results of particle spectrum in several simulations with different $\theta$.

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.95\textwidth]{fig/protons.eps} 
	\caption{Distribution of protons.}
	\label{protons}
\end{figure}
\begin{figure}[h!]
	\centering
	\includegraphics[width=0.95\textwidth]{fig/electrons.eps} 
	\caption{Distribution of electrons.}
	\label{electrons}
\end{figure}

One can see on Figures \ref{protons} and \ref{electrons} spectrum of particles with different angle $\theta$. Spectrum consists of three parts - narrow peak of cold but fast moving initial flow, wide peak of hot downstream and non-thermal accelerated component. Figures show that spectrum of accelerated particles strongly depends on angle $\theta$. If $\theta$ is less then critical value, defined by equation $c\cdot cos(\theta_{crit})=v_{shock}$, where all values are measured in upstream rest frame, particles can escape from the front and cross it several times to gain more energy. It is difficult to evaluate critical angle apriori, because we don't know compression relation of shock wave which will be created in simulation, but estimation for case of strong wave (compression relation $\sigma=4$) gets $\theta_{crit}=45^{\circ}$ in downstream frame. And it is consistent with results of our simulations.

Also, one can see that for angles less then critical, particle spectrum are higher and longer for larger $\theta$. Spectrum of nonthermal component is about 10 times higher for protons than for electrons at the same energy. It is consistent with other Particle-in-cell simulations o shock waves {\cite{Sironi2011}}.

Results show interesting effects for angles close to the critical value (black and yellow line)- electrons are still accelerating while protons are not. It can be explained by electrons accelerating in shortwave turbulence in precursor of shock wave.
